# Set up simple simulation model

This chapter is to illustrate how to use `mparseRcpp` to simulate from a simple epidemic model. To start with we introduce a simple case study system. Before we do that, load the `SimBIID` and [`outbreaks`](http://www.repidemicsconsortium.org/outbreaks/index.html) packages (the latter containing the data).

```{r, sim-simbiid, warning = F, message = F}
## load libraries
library(SimBIID)
library(outbreaks)
```

## Case Study

To illustrate some of these ideas we will use a case study of influenza in a boarding school. These data are from a paper in the BMJ in 1978 [@anonymous:1978] and provided in the [`outbreaks`](http://www.repidemicsconsortium.org/outbreaks/index.html) R package. We use a simple $SIRR_1$ model with two removal classes, $R$ and $R_1$. We will assume that the bed-rest counts in the data correspond to the number of individuals in the $R$ class, and will ignore the other time-series for the time being. This has been used various times in the literature, including @murray:2003, @devriesetal:2006 and in some of the [`pomp` tutorials](https://kingaa.github.io/sbied/mif/mif.html#an-iterated-filtering-algorithm-if2). The stochastic model we will use has event probabilities:
\begin{align*}
    P\left[S_{t + \delta t} = S_t - 1, I_{t + \delta t} = I_t + 1\right] &\approx \beta S I / N\\
    P\left[I_{t + \delta t} = I_t - 1, R_{t + \delta t} = R_t + 1\right] &\approx \gamma I\\
    P\left[R_{t + \delta t} = R_t - 1, R_{1, t + \delta t} = R_{1,t} + 1\right] &\approx \gamma_1 R
\end{align*}
The initial population size is 763 pupils, and we assume an initial introduction of infection of a single individual at day 0.

```{r, case-BS, message = F, warning = F}
## extract data from 'outbreaks'
flu <- influenza_england_1978_school
flu$day <- 1:nrow(flu)
```

`r ifelse(opts_knit$get("rmarkdown.pandoc.to") == "latex", "\\newpage", "")`

```{r, plot-BS, message = F, warning = F, fig.height = 3.5, fig.width = 3.5}
## plot data
plot(flu$day, flu$in_bed, xlab = "Day", ylab = "Number in bed-rest", type = "l")
```

## Set up simulation model

To set up the model above we need a character vector of transition rates:

```{r, sim-trans}
## set up transition rates
transitions <- c(
    "S -> beta * S * I / (S + I + R + R1) -> I", 
    "I -> gamma * I -> R",
    "R -> gamma1 * R -> R1"
)
```

We also need a vector of compartment names and parameter names:

```{r, sim-comp}
## set compartment and parameter names
compartments <- c("S", "I", "R", "R1")
pars <- c("beta", "gamma", "gamma1")
```

We then build a model using the `mparseRcpp()`:

```{r, sim-model}
## parse model
model <- mparseRcpp(
    transitions = transitions, 
    compartments = compartments,
    pars = pars
)
```

> If you wish to, you can view the parsed `Rcpp` source code by simply printing the object e.g.
> ```
> model
> ```

`r ifelse(opts_knit$get("rmarkdown.pandoc.to") == "latex", "\\newpage", "")`

The `SimBIID` package provides a `run()` method to run individual or multiple simulations from the model. By default the `mparseRcpp` creates a function that takes four arguments:

* `pars`: a vector of parameter values;
* `tstart`: the time to begin the simulation;
* `tstop`: the time to end the simulation;
* `u`: states of the system at `tstart`.

We can then pass this object to the `run()` method, along with the corresponding arguments e.g.

```{r, sim-seed, include = F}
set.seed(1)
```

```{r, sim-run}
## define parameters
simPars <- c(beta = 2.5, gamma = 1, gamma1 = 1)

## define initial states
iniStates <- c(S = 762, I = 1, R = 0, R1 = 0)

## run model
sims <- run(
    model = model,
    pars = simPars,
    tstart = 0,
    tstop = 15,
    u = iniStates
)
sims
```

Currently this outputs a `tibble` object (essentially a `data.frame`), with columns: `completed / t / u*`, where `completed` is a binary variable containing the value `1` if the epidemic finished before time point `tstop`, and `0` otherwise. Then the value `t` is the time point when the epidemic finished (or `tstop` if the epidemic was still ongoing when the function exited), and I'm using `u*` generically to correspond to a copy of the states of the system at the final time point (here $S$, $I$, $R$ and $R_1$).

Here we can see that the simulation finished at time point `t = `r round(sims$sums[3], 2)``, with the final state of the system given by ``r paste0("c(S = ", sims$sums[4], ", I = ", sims$sums[5], ", R = ", sims$sums[6], ", R1 = ", sims$sums[7], ")")``.

## Using `tspan`

Alternatively, we may want to return the states of the system at a series of pre-determined points. We can do this by adding a `tspan` argument, which is a vector of time points at which we wish to return the states. We initialise the model by telling `mparseRcpp` that we wish to include a `tspan` argument e.g.

```{r, sim-mod1}
## parse model
model <- mparseRcpp(
    transitions = transitions, 
    compartments = compartments,
    pars = pars,
    tspan = T
)
```

When we `run()` the model, we can enter a suitable `tspan` argument as a vector of time points at which we wish to return the states e.g.

```{r, sim-seed1, include = F}
set.seed(10)
```

```{r, sim-run1}
## run model
sims <- run(
    model = model,
    pars = simPars,
    tstart = 0,
    tstop = 15,
    u = iniStates,
    tspan = 1:14
)
sims
```

Here we can see that the epidemic `r ifelse(sims$sums[2] == 1, "died out", "was still going")` at time point `t = `r round(sims$sums[3], 1)``.

`r ifelse(opts_knit$get("rmarkdown.pandoc.to") == "latex", "\\newpage", "")`

## Plotting simulations

The `run()` method outputs a `SimBIID_runs` object, which can be plotted using the `plot()` function:

```{r, sim-plot, out.width = "80%"}
## plot simulation
plot(sims)
```

## Running multiple simulations

We can run multiple simulations by passing a `nrep` argument to `run()`. For example, to run 100 replicates and plot them:

```{r, sim-run2}
## run model
sims <- run(
    model = model,
    pars = simPars,
    tstart = 0,
    tstop = 15,
    u = iniStates,
    tspan = 1:14,
    nrep = 100
)
```

`r ifelse(opts_knit$get("rmarkdown.pandoc.to") == "latex", "\\newpage", "")`

```{r, sim-run21, out.width = "80%"}
sims
## plot simulations
plot(sims, quant = c(0.6, 0.7, 0.8, 0.9))
```

Here the ribbons correspond to different prediction bands, and the red line is the posterior median. We can also add individual trajectories by inputing their replicate number, for example below we add the trajectories for replicates 1 and 2 to the plot.

```{r, sim-plot1, out.width = "80%"}
## plot simulations
plot(sims, rep = c(1, 2), quant = c(0.6, 0.7, 0.8, 0.9))
```

There is an argument in `run()` to enable parallel processing if required (if the `parallel` package is installed). See help file for `run()` (e.g. `?run`) for more details.
